Studying the Performance of the Jellyfish Search Optimiser for the Application of Projection Pursuit

### 2024-06-29 optim.qmd
done edit section 5

**README**:

-   British English ("American or British usage is accepted, but not a
    mixture of these")

-   No "we" - always third person

-   Check if the affiliation information is correct

-   word use:

    -   Use jellyfish search optimiser, or JSO
    -   "hyper-parameter" rather than "hyperparameter"
    -   PP index rather than PPI, plural PP indexes
    -   datasets rather than data sets

```{r setup, echo = FALSE, message=FALSE, warning=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE)
#install the latest version of ferrn from: https://github.com/huizezhang-sherry/ferrn
library(tidyverse)
library(ggplot2)
library(patchwork)
library(ggh4x)
library(broom)
library(kableExtra)
library(ferrn)
load(here::here("data/sim_pipe_run_best.rda"))
load(here::here("data/sim_sine_6d_spline_head.rda"))
load(here::here("data/sim_sine_6d_spline_best.rda"))
load(here::here("data/sim_sine_6d_spline_projdist.rda"))
load(here::here("data/pipe_better.rda"))
load(here::here("data/pipe_jellyfish.rda"))
load(here::here("data/smoothness.rda"))
load(here::here("data/squintability.rda"))
load(here::here("data/sq_basis_dist_idx.rda"))
load(here::here("data/sim_df.rda"))
```

@@

The paper is structured as follows. @sec-background introduces the background of PP guided tour and reviews existing optimisers and index functions in the literature. @sec-theory details JSO and introduces metrics that measure different properties of PP indexes. @sec-vis outlines the data structures, numerical summaries, and visualisations used in the simulation studies conducted in @sec-app to assess JSO's performance. @sec-app presents two sets of simulation experiments, comparing JSO permformance improvements relative to an existing optimiser, and studies the impact of different PP index properties on the optimisation performance. @sec-conclusion summarises the work and provides suggestions for future directions.

@@

# Simulation details {#sec-vis}

@@

The performance of JSO is investigated in two folds: 1) compare it with a commonly used optimiser: the search-better optimiser (SBO) [@RJ-2021-105;
@laa_using_2020], and 2) examine its success rate under two hyper-parameters (number of jellyfishes and maximum number of tries). The projection pursuit problem used is finding the pipe shape using the holes index, investigated by @laa_using_2020. Fifty simulations are conducted with both JSO and the SBO, in four dimensions ($d = 6, 8, 10, 12$). JSO uses 100 jellyfishes and 100 maximum number of tries, while the SBO allows a maximum of 1000 samples at each iteration before the algorithm terminates. The results of the simulation are collected using the data structure proposed in @RJ-2021-105 is used here for assessing JSO. A sample for the sine-wave structure in 6D data using a spline index function (`splines2d`) is printed below:

@@

\newpage

# Results \[Di and Sherry\] {#sec-app}

@@

@fig-proj presents the final projections found by the two optimisers, broken down by 10th quantile, faceted by the data dimension. In the 6-dimensional data scenario, JSO consistently identifies a clear pipe shape. The SBO also finds the pipe shape but with a wide rim, suggesting a further polish search may be required. With increasing dimensions, JSO may not always identify the pipe shape due to random sampling, but it still presents a pipe shape in over 50%
of cases. When compared to the SBO JSO reaches higher index values and clearer pipe shapes across all quantiles.

@fig-proportion presents the proportion of success for each
hyper-parameter combinations. As the number of jellyfishes and maximum
tries increase, the success rate also increases. For simpler problems (6
dimensions), small parameter values (20 jellyfishes and a maximum number
of tries of 50) can already result in high success rate, while for
higher-dimensional problems (8, 10, and 12 dimensions), a combination of
100 jellyfishes and 100 maximum number of tries is necessary for at
least half of the jellyfishes to find the projection close to the optimal.

```{r}
#| fig.height = 7.5,
#| fig.width = 10,
#| label = fig-proj,
#| fig.align = "center",
#| fig.cap = "Projections found by the JSO and SBO at each 10th quantile across 50 simulations. The projection pursuit problem is to find the pipe shape using the holes index in the 6, 8, 10, and 12-dimensional spaces. The JSO uses 100 jellyfishes and a maximum number of tries of 100. The SBO uses a maximum of 1000 tries in each step of random sampling step before the algorithm terminates. In the 6-D data space, JSO always finds a clear pipe shape while the SBO also finds the pipe shape but with a wide rim. At higher data dimensions, JSO finds a higher index value and a clearer pipe shape across all the quantiles than the SBO."

@@

(p1 + ggtitle("a. JSO"))/(p2 + ggtitle("b. SBO"))
```

```{r}
#| label = fig-proportion,
#| fig.width = 6,
#| fig.height = 3,
#| fig.align = "center",
#| fig.cap = "Proportion of simulations reaches near-optimal index values in the pipe-finding problem using the holes index. The proportion is calculated based on the number of simulations, out of 50, that achieve an index value within 0.05 of the best-performing simulation. As the dimensionality increases, the proportion of simulations reaching the optimal index value increases."
pipe_sim_best <- sim_pipe_run_best |> 
  group_by(n_jellies, max_tries, d) |>
  summarise(I_best = max(I_max))

@@

<!-- The optimisation properties, including smoothness, squintability, and -->
<!-- speed, offer numerical metrics to characterise the complexity of -->
<!-- projection pursuit optimisation problems. This example investigates how -->
<!-- these metrics, along with the hyper-parameters affect the JSO -->
<!-- success rate. Simulations are conducted to obtain the success rate -->
<!-- across a wider array of projection pursuit problems, characterised by -->
<!-- shape-to-find, data dimension, and the index function, as well as -->
<!-- different hyper-parameter combinations. Smoothness and squintability are -->
<!-- calculated for each projection pursuit problem as described in -->
<!-- @sec-theory. A generalised linear model is used to construct the -->
<!-- relationship between the success rate, JSO hyper-parameters and -->
<!-- optimisation properties. -->

@tbl-smoothness-squintability presents the parameters estimated from the
Gaussian process (variance, range, smooth, and nugget) and the scaled
logistic function ($\theta_1$ to $\theta_4$) for each case. The column
"smooth" is used as the smoothness measure and the column "squint" is
calculated as equation @eq-squintability-parametric as squintability. The low
squint value of MIC and TIC index is due to its convex shape of the
index value against the projection distance, as shown in
@fig-idx-proj-dist. Comparing to other concave shapes, the maximum first
derivative happens at a smaller projection distance ($\theta_2$),
requiring the optimiser to get closer to the optimal to see a
significant change in the index value, hence a small squintability
measure.

```{r fig-idx-proj-dist}
#| fig.width = 9,
#| fig.height = 5,
#| fig.cap = "Index values versus projection distance for the 12 pipe/sine-wave finding problem, after the binning procedure for calculating the squintability measure. The index values, averaged at bin width of 0.005, are scaled from 0 to 1 for comparison (holes, TIC, and stringy). The `MIC` and `TIC` index curves are convex while others are concave. The stringy curve shows an instantaneous jump to the optimum when aproaching the best basis." 
sq_basis_dist_idx |>
  ggplot(aes(x = dist, y = index, group = index_name)) +
  geom_line() + 
  facet_nested_wrap(~index_name + n, nrow = 3, labeller = label_both) + 
  xlab("projection distance") + ylab("index value") 
```

@@

```{r}
#| label: tbl-mod-data
#| tbl-cap: The first 7 rows of the datasets processed for modelling. 
sim_df |> 
  dplyr::select(index, d, smoothness, squintability, n_jellies, max_tries, 
                -I_max, P_J, time) |> 
  arrange(index, d, n_jellies, max_tries) |> 
  head(7) |> 
  kable(digits = 3,
        col.names = c("Index", "D", "Smoothness", "Squintability", 
                      "Num. of jellyfish", "Max. Num. of tries", 
                      "Prop. of success", "Time"), 
        booktabs = T) |> 
  kable_styling(latex_options = "hold_position", font_size = 7)
```

```{r}
#| label: tbl-mod-output
#| tbl-cap: Model estimates of proportion of jellyfish success on optimisation properties and jellyfish hyper-parameters from the generalised linear model with a binomial family and a logit link function. The variable smoothness, squintability, number of jellyfish and maximum number of tries are positively associated with JSO success rate while data dimension and being flagged as long runtime are negatively associated with the success rate. The variable squintability and dimension are significant, suggesting their importance relative to jellyfish hyper-parameters in the optimisation success.
sim_df2 <- sim_df |>   
  mutate(n_jellies = n_jellies/10,
         max_tries = max_tries/10,
         long_time = ifelse(time > 20, 1, 0),
         P_J = ifelse(index == "stringy", 0, P_J))

mod1 <- glm(P_J ~ smoothness + squintability + d + long_time + n_jellies + max_tries, data = sim_df2, family = binomial)
broom::tidy(mod1) |> 
  mutate(term = c("Intercept", "Smoothness", "Squintability", "dimension (d)", 
                  "long time", "number of jellyfish", "maximum number of tries")) |>
  kable(digits = 3) |> 
  column_spec(c(1,5), border_right = TRUE) |> 
  column_spec(1, border_left = TRUE) |> 
  kable_styling(font_size = 7) 
```

@tbl-mod-data combines all the variables calculated above (JSO
success rate, hyper-parameters, and optimisation properties) into
one table. @tbl-mod-output presents the model
outputs. The model suggests that JSO success rate is positively
associated with the two hyper-parameters, as well as with the
optimisation properties -- smoothness and squintability. Specifically,
using 10 more jellyfishes and 10 more tries will increase the odd ratio of
success by `r round(100 * (exp(broom::tidy(mod1)$estimate[6])-1), 2)`%
and `r round(100 * (exp(broom::tidy(mod1)$estimate[7])-1), 2)`%,
respectively. However, being flagged with long runtime and an increase
of data dimension will reduce the success rate by
`r round(exp(broom::tidy(mod1)$estimate[5]) * 100, 2)`% and
`r round(exp(broom::tidy(mod1)$estimate[4]) * 100, 2)`%, respectively.
The variable `squintability` and `dimension` are significant, suggesting
their importance relative to JSO hyper-parameters in the optimisation
success. Users should consider increasing the number of jellyfish and the number of
tries to improve the success rate, until JSO exceeds a 30-second
runtime.

@@

>>>>

```{r setup, echo = FALSE, message=FALSE, warning=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE)
library(tidyverse)
library(ggplot2)
library(patchwork)
library(ggh4x)
library(broom)
library(kableExtra)
library(ferrn)
load(here::here("data/sim_pipe_run_best.rda"))
load(here::here("data/sim_sine_6d_spline_head.rda"))
load(here::here("data/sim_sine_6d_spline_best.rda"))
load(here::here("data/sim_sine_6d_spline_projdist.rda"))
load(here::here("data/pipe_better.rda"))
load(here::here("data/pipe_jellyfish.rda"))
load(here::here("data/smoothness.rda"))
load(here::here("data/squintability.rda"))
load(here::here("data/sq_basis_dist_idx.rda"))
load(here::here("data/sim_df.rda"))
```

@@

The paper is structured as follows. @sec-background introduces the background of PP guided tour and reviews existing optimisers and index functions in the literature. @sec-theory details JSO and introduces metrics that measure different properties of PP indexes. @sec-sim-deets outlines the data structures, numerical summaries, and visualisations used in the simulation studies conducted in @sec-app to assess JSO's performance. @sec-sim-res presents two sets of simulation experiments, comparing JSO permformance improvements relative to an existing optimiser, and studies the impact of different PP index properties on the optimisation performance. @sec-conclusion summarises the work and provides suggestions for future directions.

@@

# Simulation details {#sec-sim-deets}

@@

The performance of JSO is investigated in two folds: 1) compare it with a commonly used optimiser: the Csreeping Random Search (CRS) [@RJ-2021-105;
@laa_using_2020], and 2) examine its success rate under two hyper-parameters (number of jellyfishes and maximum number of tries). The projection pursuit problem used is finding the pipe shape using the holes index, investigated by @laa_using_2020. Fifty simulations are conducted with both JSO and CRS, in four dimensions ($d = 6, 8, 10, 12$). JSO uses 100 jellyfishes and 100 maximum number of tries, while the CRS allows a maximum of 1000 samples at each iteration before the algorithm terminates. The results of the simulation are collected using the data structure proposed in @RJ-2021-105 is used here for assessing JSO. A sample for the sine-wave structure in 6D data using a spline index function (`splines2d`) is printed below:

@@

# Results \[Di and Sherry\] {#sec-sim-res}

The results from the two simulations described in @sec-sim-deets are analysed based on the final projections across two optimisers (JSO and CRS) and the success rate across JSO hyper-parameters. Smoothness and squintability are calculated across a collection of pipe-finding and sine-wave finding problems to construct the relationship between success rate, JSO hyper-parameters, and optimisation properties.

@@

The final projections found by the two optimisers (JSO and CRS) are presented in 
@fig-proj, broken down by 10th quantile, faceted by the data dimension. The JSO uses 100 jellyfishes and a maximum number of tries of 100. The CRS uses a maximum of 1000 tries in each step of random sampling step before the algorithm terminates. In the 6-dimensional data scenario, JSO consistently identifies a clear pipe shape. The CRS also finds the pipe shape but with a wide rim, suggesting a further polish search may be required. With increasing dimensions, JSO may not always identify the pipe shape due to random sampling, but it still presents a pipe shape in over 50% of cases. When compared to the CRS, JSO reaches higher index values and clearer pipe shapes across all quantiles. This suggests JSO can be more effective when searching higher dimension spaces, which may be preferred by users.

The success rate calculated for each hyper-parameter combination is presented in @fig-proportion.
As the number of jellyfishes and maximum tries increase, the success rate also increases. For simpler problems (6 dimensions), small parameter values (20 jellyfishes and a maximum
of 50 tries) can achieve a high success rate. However, for higher-dimensional problems (8, 10, and 12 dimensions), larger parameter values (i.e. 100 jellyfishes and a maximum of 100 tries) are necessary. Increasing both parameters enhances the performance of JSO, but it also extends the computational time required for the optimisation, which can be computationally intensive when evaluating the index function (such as scagnostic indexes) multiple times across numerous iterations.

```{r}
#| fig.height = 7.5,
#| fig.width = 10,
#| label = fig-proj,
#| fig.align = "center",
#| fig.cap = "Projections found by the JSO and CRS at each 10th quantile across 50 simulations. The projection pursuit problem is to find the pipe shape using the holes index in the 6, 8, 10, and 12-dimensional spaces. The JSO uses 100 jellyfishes and a maximum number of tries of 100. The CRS uses a maximum of 1000 tries in each step of random sampling step before the algorithm terminates. In the 6-D data space, JSO always finds a clear pipe shape while the CRS also finds the pipe shape but with a wide rim. At higher data dimensions, JSO finds a higher index value and a clearer pipe shape across all the quantiles than the CRS"

@@

(p1 + ggtitle("a. JSO"))/(p2 + ggtitle("b. CRS"))
```

```{r}
#| label = fig-proportion,
#| fig.width = 6,
#| fig.height = 2.5,
#| fig.align = "center",
#| fig.cap = "Proportion of simulations reaches near-optimal index values in the pipe-finding problem using the holes index. The proportion is calculated based on the number of simulations, out of 50, that achieve an index value within 0.05 of the best-performing simulation. As the dimensionality increases, the proportion of simulations reaching the optimal index value increases."
pipe_sim_best <- sim_pipe_run_best |> 
  group_by(n_jellies, max_tries, d) |>
  summarise(I_best = max(I_max))

@@

```{r}
sim_df2 <- sim_df |>   
  mutate(n_jellies = n_jellies/10,
         max_tries = max_tries/10,
         long_time = ifelse(time > 20, 1, 0),
         P_J = ifelse(index == "stringy", 0, P_J))

mod1 <- glm(P_J ~ smoothness + squintability + d + long_time + n_jellies + max_tries, data = sim_df2, family = binomial)
```

<!-- The low -->
<!-- squint value of MIC and TIC index is due to its convex shape of the -->
<!-- index value against the projection distance, as shown in -->
<!-- @fig-idx-proj-dist. Comparing to other concave shapes, the maximum first -->
<!-- derivative happens at a smaller projection distance ($\theta_2$), -->
<!-- requiring the optimiser to get closer to the optimal to see a -->
<!-- significant change in the index value, hence a small squintability -->
<!-- measure.  -->

The optimisation properties, including smoothness and squintability,
offer numerical metrics to characterise the complexity of
projection pursuit optimisation problems. @tbl-smoothness-squintability presents 
the parameters estimated from the
Gaussian process (variance, range, smooth, and nugget) and the scaled
logistic function ($\theta_1$ to $\theta_4$) for calculating both metrics for each case considered in the simulation. The column
"smooth" is used as the smoothness metric and the column "squint" is
calculated as equation @eq-squintability-parametric as the squintability metric.  @tbl-mod-output presents the results of fitting a generalised linear model with a binomial family and a logit link
function to the data in @tbl-mod-data, where all the three components (success rate, JS hyper-parameters, and optimisation properties) are combined. The model suggests that JSO success rate is positively
associated with the two hyper-parameters, as well as with the
optimisation properties: smoothness and squintability. Specifically,
using 10 more jellyfishes and 10 more tries increases the odd ratio of
success by `r round(100 * (exp(broom::tidy(mod1)$estimate[6])-1), 2)`%
and `r round(100 * (exp(broom::tidy(mod1)$estimate[7])-1), 2)`%,
respectively. However, being flagged with long runtime and an increase
of data dimension reduce the success rate by
`r round(exp(broom::tidy(mod1)$estimate[5]) * 100, 2)`% and
`r round(exp(broom::tidy(mod1)$estimate[4]) * 100, 2)`%, respectively.
The variable `squintability` and `dimension` are significant, suggesting
their importance relative to JSO hyper-parameters in the optimisation
success. In defining a projection pursuit problem, factors such as the shape-to-find, the index function used, and the data dimension, determine the optimisation properties like smoothness and squintability. These metrics can then be compared across different problems to understand their relative complexity. The results suggest that squintability has a more significant influence than smoothness on the success rate of JSO. Once the characteristics of the projection pursuit problem are fully understood, increasing JSO hyper-parameters can enhance the search effectiveness, however, it is also important to consider the resulting increase in computational complexity.

```{r fig-idx-proj-dist, eval = FALSE}
#| fig.width = 9,
#| fig.height = 5,
#| fig.cap = "Index values versus projection distance for the 12 pipe/sine-wave finding problem, after the binning procedure for calculating the squintability measure. The index values, averaged at bin width of 0.005, are scaled from 0 to 1 for comparison (holes, TIC, and stringy). The `MIC` and `TIC` index curves are convex while others are concave. The stringy curve shows an instantaneous jump to the optimum when aproaching the best basis." 
sq_basis_dist_idx |>
  ggplot(aes(x = dist, y = index, group = index_name)) +
  geom_line() + 
  facet_nested_wrap(~index_name + n, nrow = 3, labeller = label_both) + 
  xlab("projection distance") + ylab("index value") +
  theme_bw()
```

@@

```{r}
#| label: tbl-mod-data
#| tbl-cap: The first 7 rows of the datasets processed for modelling. 
sim_df |> 
  dplyr::select(index, d, smoothness, squintability, n_jellies, max_tries, 
                -I_max, P_J, time) |> 
  arrange(index, d, n_jellies, max_tries) |> 
  head(7) |> 
  kable(digits = 3,
        col.names = c("Index", "D", "Smoothness", "Squintability", 
                      "Num. of jellyfish", "Max. Num. of tries", 
                      "Success rate", "Time"), 
        booktabs = T) |> 
  kable_styling(latex_options = "hold_position", font_size = 7)
```

```{r}
#| label: tbl-mod-output
#| tbl-cap: Model estimates of proportion of jellyfish success on optimisation properties and jellyfish hyper-parameters from the generalised linear model with a binomial family and a logit link function. The variable smoothness, squintability, number of jellyfish and maximum number of tries are positively associated with JSO success rate while data dimension and being flagged as long runtime are negatively associated with the success rate. The variable squintability and dimension are significant, suggesting their importance relative to jellyfish hyper-parameters in the optimisation success.
broom::tidy(mod1) |> 
  mutate(term = c("Intercept", "Smoothness", "Squintability", "dimension (d)", 
                  "long time", "number of jellyfish", "maximum number of tries")) |>
  kable(digits = 3) |> 
  column_spec(c(1,5), border_right = TRUE) |> 
  column_spec(1, border_left = TRUE) |> 
  kable_styling(font_size = 7) 
```

@@

Main findings:

  - JSO tends to obtian a clearer projection and higher index value when searching higher dimension spaces than CRS (and other current optimisers) - Figure 4

  - Increasing JS hyper-parameter enhances its performance, but the computational time required for the optimisation can quickly become intensive, especially when evaluating the index function (such as scagnostic indexes) multiple times across numerous iterations. - Figure 5

  - Smoothness and squintability can be calculated and compared across different problems to understand their relative complexity. Once a projection pursuit problem is fully characterised (by the shape-to-find, the index function used, and the data dimension), increasing JSO hyper-parameters can enhance the search effectiveness, however, the resulting increase in computational complexity will also need to be considered. - modelling

### 2024-06-29 optim.tex
done edit section 5

\textbf{README}:

\begin{itemize}
\item
  British English (``American or British usage is accepted, but not a
  mixture of these'')
\item
  No ``we'' - always third person
\item
  Check if the affiliation information is correct
\item
  word use:

  \begin{itemize}
  \tightlist
  \item
    Use jellyfish search optimiser, or JSO
  \item
    ``hyper-parameter'' rather than ``hyperparameter''
  \item
    PP index rather than PPI, plural PP indexes
  \item
    datasets rather than data sets
  \end{itemize}
\end{itemize}

@@

The paper is structured as follows. Section~\ref{sec-background}
introduces the background of PP guided tour and reviews existing
optimisers and index functions in the literature.
Section~\ref{sec-theory} details JSO and introduces metrics that measure
different properties of PP indexes. Section~\ref{sec-vis} outlines the
data structures, numerical summaries, and visualisations used in the
simulation studies conducted in Section~\ref{sec-app} to assess JSO's
performance. Section~\ref{sec-app} presents two sets of simulation
experiments, comparing JSO permformance improvements relative to an
existing optimiser, and studies the impact of different PP index
properties on the optimisation performance. Section~\ref{sec-conclusion}
summarises the work and provides suggestions for future directions.

@@

\begin{tcolorbox}[enhanced jigsaw, colframe=quarto-callout-note-color-frame, bottomtitle=1mm, bottomrule=.15mm, breakable, left=2mm, coltitle=black, colback=white, opacitybacktitle=0.6, colbacktitle=quarto-callout-note-color!10!white, toptitle=1mm, leftrule=.75mm, title={Algorithm: Jellyfish Optimizer Pseudo Code}, rightrule=.15mm, titlerule=0mm, arc=.35mm, toprule=.15mm, opacityback=0]

@@

\section{Simulation details}\label{sec-vis}

@@

The performance of JSO is investigated in two folds: 1) compare it with
a commonly used optimiser: the search-better optimiser (SBO)
\citep{RJ-2021-105, laa_using_2020}, and 2) examine its success rate
under two hyper-parameters (number of jellyfishes and maximum number of
tries). The projection pursuit problem used is finding the pipe shape
using the holes index, investigated by \citet{laa_using_2020}. Fifty
simulations are conducted with both JSO and the SBO, in four dimensions
(\(d = 6, 8, 10, 12\)). JSO uses 100 jellyfishes and 100 maximum number
of tries, while the SBO allows a maximum of 1000 samples at each
iteration before the algorithm terminates. The results of the simulation
are collected using the data structure proposed in \citet{RJ-2021-105}
is used here for assessing JSO. A sample for the sine-wave structure in
6D data using a spline index function (\texttt{splines2d}) is printed
below:

@@

\newpage

\section{Results {[}Di and Sherry{]}}\label{sec-app}

@@

Figure~\ref{fig-proj} presents the final projections found by the two
optimisers, broken down by 10th quantile, faceted by the data dimension.
In the 6-dimensional data scenario, JSO consistently identifies a clear
pipe shape. The SBO also finds the pipe shape but with a wide rim,
suggesting a further polish search may be required. With increasing
dimensions, JSO may not always identify the pipe shape due to random
sampling, but it still presents a pipe shape in over 50\% of cases. When
compared to the SBO JSO reaches higher index values and clearer pipe
shapes across all quantiles.

Figure~\ref{fig-proportion} presents the proportion of success for each
hyper-parameter combinations. As the number of jellyfishes and maximum
tries increase, the success rate also increases. For simpler problems (6
dimensions), small parameter values (20 jellyfishes and a maximum number
of tries of 50) can already result in high success rate, while for
higher-dimensional problems (8, 10, and 12 dimensions), a combination of
100 jellyfishes and 100 maximum number of tries is necessary for at
least half of the jellyfishes to find the projection close to the
optimal.

@@

\caption{\label{fig-proj}Projections found by the JSO and SBO at each
10th quantile across 50 simulations. The projection pursuit problem is
to find the pipe shape using the holes index in the 6, 8, 10, and
12-dimensional spaces. The JSO uses 100 jellyfishes and a maximum number
of tries of 100. The SBO uses a maximum of 1000 tries in each step of
random sampling step before the algorithm terminates. In the 6-D data
space, JSO always finds a clear pipe shape while the SBO also finds the
pipe shape but with a wide rim. At higher data dimensions, JSO finds a
higher index value and a clearer pipe shape across all the quantiles
than the SBO.}

@@

Table~\ref{tbl-smoothness-squintability} presents the parameters
estimated from the Gaussian process (variance, range, smooth, and
nugget) and the scaled logistic function (\(\theta_1\) to \(\theta_4\))
for each case. The column ``smooth'' is used as the smoothness measure
and the column ``squint'' is calculated as equation
\ref{eq-squintability-parametric} as squintability. The low squint value
of MIC and TIC index is due to its convex shape of the index value
against the projection distance, as shown in
Figure~\ref{fig-idx-proj-dist}. Comparing to other concave shapes, the
maximum first derivative happens at a smaller projection distance
(\(\theta_2\)), requiring the optimiser to get closer to the optimal to
see a significant change in the index value, hence a small squintability
measure.

\includegraphics{optim_files/figure-pdf/fig-idx-proj-dist-1.pdf}

\caption{\label{fig-idx-proj-dist}Index values versus projection
distance for the 12 pipe/sine-wave finding problem, after the binning
procedure for calculating the squintability measure. The index values,
averaged at bin width of 0.005, are scaled from 0 to 1 for comparison
(holes, TIC, and stringy). The \texttt{MIC} and \texttt{TIC} index
curves are convex while others are concave. The stringy curve shows an
instantaneous jump to the optimum when aproaching the best basis.}

@@

\toprule
Index & D & Smoothness & Squintability & Num. of jellyfish & Max. Num. of tries & Prop. of success & Time\\
\midrule
MIC & 6 & 2.457 & 1.257 & 20 & 50 & 0.12 & 2.479 secs\\
MIC & 6 & 2.457 & 1.257 & 20 & 100 & 0.24 & 8.950 secs\\
MIC & 6 & 2.457 & 1.257 & 50 & 50 & 0.52 & 5.651 secs\\
MIC & 6 & 2.457 & 1.257 & 50 & 100 & 0.64 & 13.223 secs\\
MIC & 6 & 2.457 & 1.257 & 100 & 50 & 0.76 & 19.453 secs\\
\addlinespace
MIC & 8 & 2.636 & 0.772 & 20 & 50 & 0.08 & 2.566 secs\\
MIC & 8 & 2.636 & 0.772 & 20 & 100 & 0.08 & 4.960 secs\\
\bottomrule

@@

Table~\ref{tbl-mod-data} combines all the variables calculated above
(JSO success rate, hyper-parameters, and optimisation properties) into
one table. Table~\ref{tbl-mod-output} presents the model outputs. The
model suggests that JSO success rate is positively associated with the
two hyper-parameters, as well as with the optimisation properties --
smoothness and squintability. Specifically, using 10 more jellyfishes
and 10 more tries will increase the odd ratio of success by 24.11\% and
11.93\%, respectively. However, being flagged with long runtime and an
increase of data dimension will reduce the success rate by 41.72\% and
53.36\%, respectively. The variable \texttt{squintability} and
\texttt{dimension} are significant, suggesting their importance relative
to JSO hyper-parameters in the optimisation success. Users should
consider increasing the number of jellyfish and the number of tries to
improve the success rate, until JSO exceeds a 30-second runtime.

@@

>>>>

@@

The paper is structured as follows. Section~\ref{sec-background}
introduces the background of PP guided tour and reviews existing
optimisers and index functions in the literature.
Section~\ref{sec-theory} details JSO and introduces metrics that measure
different properties of PP indexes. Section~\ref{sec-sim-deets} outlines
the data structures, numerical summaries, and visualisations used in the
simulation studies conducted in \textbf{?@sec-app} to assess JSO's
performance. Section~\ref{sec-sim-res} presents two sets of simulation
experiments, comparing JSO permformance improvements relative to an
existing optimiser, and studies the impact of different PP index
properties on the optimisation performance. Section~\ref{sec-conclusion}
summarises the work and provides suggestions for future directions.

@@

\begin{tcolorbox}[enhanced jigsaw, opacitybacktitle=0.6, colbacktitle=quarto-callout-note-color!10!white, breakable, opacityback=0, toptitle=1mm, leftrule=.75mm, titlerule=0mm, coltitle=black, colframe=quarto-callout-note-color-frame, bottomrule=.15mm, rightrule=.15mm, arc=.35mm, toprule=.15mm, colback=white, bottomtitle=1mm, title={Algorithm: Jellyfish Optimizer Pseudo Code}, left=2mm]

@@

\section{Simulation details}\label{sec-sim-deets}

@@

The performance of JSO is investigated in two folds: 1) compare it with
a commonly used optimiser: the Csreeping Random Search (CRS)
\citep{RJ-2021-105, laa_using_2020}, and 2) examine its success rate
under two hyper-parameters (number of jellyfishes and maximum number of
tries). The projection pursuit problem used is finding the pipe shape
using the holes index, investigated by \citet{laa_using_2020}. Fifty
simulations are conducted with both JSO and CRS, in four dimensions
(\(d = 6, 8, 10, 12\)). JSO uses 100 jellyfishes and 100 maximum number
of tries, while the CRS allows a maximum of 1000 samples at each
iteration before the algorithm terminates. The results of the simulation
are collected using the data structure proposed in \citet{RJ-2021-105}
is used here for assessing JSO. A sample for the sine-wave structure in
6D data using a spline index function (\texttt{splines2d}) is printed
below:

@@

\section{Results {[}Di and Sherry{]}}\label{sec-sim-res}

The results from the two simulations described in
Section~\ref{sec-sim-deets} are analysed based on the final projections
across two optimisers (JSO and CRS) and the success rate across JSO
hyper-parameters. Smoothness and squintability are calculated across a
collection of pipe-finding and sine-wave finding problems to construct
the relationship between success rate, JSO hyper-parameters, and
optimisation properties.

@@

The final projections found by the two optimisers (JSO and CRS) are
presented in Figure~\ref{fig-proj}, broken down by 10th quantile,
faceted by the data dimension. The JSO uses 100 jellyfishes and a
maximum number of tries of 100. The CRS uses a maximum of 1000 tries in
each step of random sampling step before the algorithm terminates. In
the 6-dimensional data scenario, JSO consistently identifies a clear
pipe shape. The CRS also finds the pipe shape but with a wide rim,
suggesting a further polish search may be required. With increasing
dimensions, JSO may not always identify the pipe shape due to random
sampling, but it still presents a pipe shape in over 50\% of cases. When
compared to the CRS, JSO reaches higher index values and clearer pipe
shapes across all quantiles. This suggests JSO can be more effective
when searching higher dimension spaces, which may be preferred by users.

The success rate calculated for each hyper-parameter combination is
presented in Figure~\ref{fig-proportion}. As the number of jellyfishes
and maximum tries increase, the success rate also increases. For simpler
problems (6 dimensions), small parameter values (20 jellyfishes and a
maximum of 50 tries) can achieve a high success rate. However, for
higher-dimensional problems (8, 10, and 12 dimensions), larger parameter
values (i.e.~100 jellyfishes and a maximum of 100 tries) are necessary.
Increasing both parameters enhances the performance of JSO, but it also
extends the computational time required for the optimisation, which can
be computationally intensive when evaluating the index function (such as
scagnostic indexes) multiple times across numerous iterations.

@@

\caption{\label{fig-proj}Projections found by the JSO and CRS at each
10th quantile across 50 simulations. The projection pursuit problem is
to find the pipe shape using the holes index in the 6, 8, 10, and
12-dimensional spaces. The JSO uses 100 jellyfishes and a maximum number
of tries of 100. The CRS uses a maximum of 1000 tries in each step of
random sampling step before the algorithm terminates. In the 6-D data
space, JSO always finds a clear pipe shape while the CRS also finds the
pipe shape but with a wide rim. At higher data dimensions, JSO finds a
higher index value and a clearer pipe shape across all the quantiles
than the CRS}

@@

The optimisation properties, including smoothness and squintability,
offer numerical metrics to characterise the complexity of projection
pursuit optimisation problems. Table~\ref{tbl-smoothness-squintability}
presents the parameters estimated from the Gaussian process (variance,
range, smooth, and nugget) and the scaled logistic function
(\(\theta_1\) to \(\theta_4\)) for calculating both metrics for each
case considered in the simulation. The column ``smooth'' is used as the
smoothness metric and the column ``squint'' is calculated as equation
\ref{eq-squintability-parametric} as the squintability metric.
Table~\ref{tbl-mod-output} presents the results of fitting a generalised
linear model with a binomial family and a logit link function to the
data in Table~\ref{tbl-mod-data}, where all the three components
(success rate, JS hyper-parameters, and optimisation properties) are
combined. The model suggests that JSO success rate is positively
associated with the two hyper-parameters, as well as with the
optimisation properties: smoothness and squintability. Specifically,
using 10 more jellyfishes and 10 more tries increases the odd ratio of
success by 24.11\% and 11.93\%, respectively. However, being flagged
with long runtime and an increase of data dimension reduce the success
rate by 41.72\% and 53.36\%, respectively. The variable
\texttt{squintability} and \texttt{dimension} are significant,
suggesting their importance relative to JSO hyper-parameters in the
optimisation success. In defining a projection pursuit problem, factors
such as the shape-to-find, the index function used, and the data
dimension, determine the optimisation properties like smoothness and
squintability. These metrics can then be compared across different
problems to understand their relative complexity. The results suggest
that squintability has a more significant influence than smoothness on
the success rate of JSO. Once the characteristics of the projection
pursuit problem are fully understood, increasing JSO hyper-parameters
can enhance the search effectiveness, however, it is also important to
consider the resulting increase in computational complexity.

@@

\toprule
Index & D & Smoothness & Squintability & Num. of jellyfish & Max. Num. of tries & Success rate & Time\\
\midrule
MIC & 6 & 2.457 & 1.257 & 20 & 50 & 0.12 & 2.479 secs\\
MIC & 6 & 2.457 & 1.257 & 20 & 100 & 0.24 & 8.950 secs\\
MIC & 6 & 2.457 & 1.257 & 50 & 50 & 0.52 & 5.651 secs\\
MIC & 6 & 2.457 & 1.257 & 50 & 100 & 0.64 & 13.223 secs\\
MIC & 6 & 2.457 & 1.257 & 100 & 50 & 0.76 & 19.453 secs\\
\addlinespace
MIC & 8 & 2.636 & 0.772 & 20 & 50 & 0.08 & 2.566 secs\\
MIC & 8 & 2.636 & 0.772 & 20 & 100 & 0.08 & 4.960 secs\\
\bottomrule

@@

@@

Main findings:

\begin{itemize}
\item
  JSO tends to obtian a clearer projection and higher index value when
  searching higher dimension spaces than CRS (and other current
  optimisers) - Figure 4
\item
  Increasing JS hyper-parameter enhances its performance, but the
  computational time required for the optimisation can quickly become
  intensive, especially when evaluating the index function (such as
  scagnostic indexes) multiple times across numerous iterations. -
  Figure 5
\item
  Smoothness and squintability can be calculated and compared across
  different problems to understand their relative complexity. Once a
  projection pursuit problem is fully characterised (by the
  shape-to-find, the index function used, and the data dimension),
  increasing JSO hyper-parameters can enhance the search effectiveness,
  however, the resulting increase in computational complexity will also
  need to be considered. - modelling
\end{itemize}
